Type-Safe Modular Parsing

### 2016-12-28 paper/sections/Introduction.tex
some text

To address the modularity limitations of Programming Languages several
different approaches have been proposed in the past. Existing
approaches can be broadly divided into two categories:
\emph{syntactic} or \emph{semantic} modularization
techniques. Syntactic modularization techniques are quite popular in
practice. Examples include many tools for developing Software-Product
Lines~\cite{}, some Language Workbenches~\cite{}, or extensible parser
generators~\cite{}.  Most syntactic approaches employ textual
composition techniques such as \emph{superimposition}~\cite{} to
enable the development modular program features. Such textual
composition techniques collect the code for multiple features and
merge it together when a concrete combination of features is needed
for a particular program. The typical drawback of such techniques is
the lack of modular type-checking and separate compilation:
``\emph{most feature-oriented implementation mechanisms lack proper
  interfaces and support neither modular type checking nor separate
  compilation}''~\cite{kastner11road}\bruno{reference to ``The road to
  Feature Modularity''}. Syntactic modularization techniques have also
been applied to the problem of \emph{extensible parsing}. There are
several approaches~\cite{} that enable the development of
\emph{syntactically} modular parsers or grammars. However these
approaches do not support separate compilation or modular
type-checking either.\bruno{mention that a reason why syntactic 
modularization techniques are popular is simplicity (in implementation
and also use).}

Semantic modularization techniques move away
from syntactic composition techniques that rely on the textual source
code. This allows us to go one step further in terms of modularity,
and also enable components or features to be modularly type-checked
and separately compiled. Modular type-checking and separate
compilation are desirable properties to have from a Software
Engineering point-of-view, and enable the composition of compiled
binaries as well as ensuring the type-safety of the code composed of
multiple components. Examples of semantic modularization techniques 
include various approaches to \emph{family polymorphism}~\cite{}, as 
well as several techniques for solving the Expression Problem. 
\bruno{strenghten? mention dynamic languages (separate compilation 
but no modular type-checking), and techniques that weaken type-safety 
requirements.} Most semantic
modularization techniques have focused on operations that traverse or 
or process extensible datastructures, such as ASTs.
\bruno{Object Algebras here?} However, as far as
we know there is little work on operations that build/produce ASTs. 
In particular the problem of modular parsing has not been studied in
semantic modularization approaches. This is a shame because, to
realise the vision of software product-lines for programming
languages, modular parsing is a necessity.

>>>>

To address the modularity limitations of Programming Languages several
different approaches have been proposed in the past. Existing
approaches can be broadly divided into two categories:
\emph{syntactic} or \emph{semantic} modularization
techniques. Syntactic modularization techniques are quite popular in
practice, due to their simplicity of implementation and use. 
Examples include many tools for developing Software-Product
Lines~\cite{}, some Language Workbenches~\cite{}, or extensible parser
generators~\cite{}.  Most syntactic approaches employ textual
composition techniques such as \emph{superimposition}~\cite{} to
enable the development modular program features. Such textual
composition techniques collect the code for multiple features and
merge it together when a concrete combination of features is needed
for a particular program. The typical drawback of such techniques is
the lack of modular type-checking and separate compilation:
``\emph{most feature-oriented implementation mechanisms lack proper
  interfaces and support neither modular type checking nor separate
  compilation}''~\cite{kastner11road}\bruno{reference to ``The road to
  Feature Modularity''}. Syntactic modularization techniques have also
been applied to the problem of \emph{extensible parsing}. There are
several approaches~\cite{} that enable the development of
\emph{syntactically} modular parsers or grammars. However these
approaches do not support separate compilation or modular
type-checking either.

Semantic modularization techniques go one step further in terms of modularity,
and also enable components or features to be modularly type-checked
and separately compiled. Modular type-checking and separate
compilation are desirable properties to have from a software
engineering point-of-view. Furthermore they enable the composition of compiled
binaries as well as ensuring the type-safety of the code composed of
multiple components. Examples of semantic modularization techniques 
include various approaches to \emph{family polymorphism}~\cite{},
virtual classes~\cite{}, as 
well as several other techniques for solving the Expression Problem~\cite{}. 
\bruno{strenghten? mention dynamic languages (separate compilation 
but no modular type-checking), and techniques that weaken type-safety 
requirements.}

Semantic modularization techniques are less widely used in practice 
than syntactic techniques. This is partly due to the perceived need
for sophisticated type systems, which are not available in mainstream 
languages and may require more knowledge from users. However, recently 
several lightweight modularization techniques have been shown to work 
in mainstream programming languages like Java or Scala. Object
Algebras are one such technique, which 
Most semantic
modularization techniques have focused on operations that traverse or 
or process extensible datastructures, such as ASTs.
\bruno{Object Algebras here?} However, as far as
we know there is little work on operations that build/produce ASTs. 
In particular the problem of modular parsing has not been studied in
semantic modularization approaches. This is a shame because, to
realise the vision of software product-lines for programming
languages, modular parsing is a necessity.
